---
title: "DEXSeq Isoform Analysis"
format: html
---

### Load Libraries

```{r}
#| message: false
library(DEXSeq)
library(tidyverse)
library(dplyr)
library(msigdbr)
library(fgsea)
```

```{r}
source("99_proj_func.R")
```

# Ocular Motor Neurons

```{r}
isoCm_Ocular <- read.csv("data/OcularIsoformCountMatrix.csv", row.names = 1)
Ocular_Metadata <- read.csv("data/OcularSampleMetadata.csv", row.names = 1)
IsoformInfo <- read.csv("data/isoformInfo.csv")
```

```{r}
OcularResult <- dexseq_analysis(isoCm_Ocular, Ocular_Metadata, IsoformInfo)
```

### Compare to Differential Expression:

```{r}
alpha <- 0.05  # threshold

OcularResultGene <- OcularResult |> 
  as.data.frame() |> 
  as_tibble() |>
  filter(!is.na(pvalue)) |> 
  select(
    gene_id = groupID,
    isoform_id = featureID,
    stat,
    log2FC = log2fold_SOD1_C9ORF72,
    pvalue,
    padj
  ) |> 
  group_by(gene_id) |> 
  slice_min(
    order_by = pvalue,
    n = 1,
    with_ties = FALSE
  )  |> 
  ungroup()  |> 
  mutate(
    padj = p.adjust(pvalue, method = 'fdr')
  ) |>
  filter(!is.na(padj))

head(OcularResultGene)
```

```{r}
dex_gene <- OcularResultGene |> 
  mutate(gene_id,
         dex_padj = padj,
         dex_sig = !is.na(padj) & padj < alpha)
head(dex_gene)
```

```{r}
load("data/Ocular_dseqResult.RData")
OcularDesq <- res_ocular
OcularDesq$padj <- as.numeric(OcularDesq$padj)
```

```{r}
OcularDesq <- OcularDesq |>
  as.data.frame() |> 
  rownames_to_column("gene_id") |>
  mutate(deseq_sig = !is.na(padj) & padj < alpha) |>
  filter(!is.na(padj))
head(OcularDesq)
```

```{r}
# Gene universe = genes present in BOTH results
univ <- intersect(dex_gene$gene_id, OcularDesq$gene_id)
both <- dex_gene |> 
  filter(gene_id %in% univ) |> 
  mutate(gene_id, dex_sig) |> 
  inner_join(OcularDesq |> 
                filter(gene_id %in% univ) |> 
                mutate(gene_id, deseq_sig),
             by = "gene_id")

contig_tab <- table(DEXSeq = both$dex_sig, DESeq2 = both$deseq_sig)
contig_tab
```

**Fisher Test**

```{r}
ft <- fisher.test(contig_tab, alternative = "greater")
ft
```

-   There is **very strong overlap** between genes detected by DEXSeq and DESeq2.

-   When DEXSeq identifies a gene with differential isoform usage, that gene is **much more likely than random** also to show differential expression at the gene level.

-   Power is different across methods—DEXSeq finds many more significant genes (isoform-level testing is more sensitive), but the ones shared with DESeq2 are **highly enriched**.

### Compare to Differential Expression at Gene-set Level

```{r}
BP_df = msigdbr(species = "human", category = "C5", subcategory = "BP")
BP_list = split(BP_df$ensembl_gene, BP_df$gs_name)
```

```{r}
deseq_names <- sub("\\.\\d+$", "", OcularDesq$gene_id)
deseq_vals <- OcularDesq$stat
stats_deseq <- setNames(deseq_vals, deseq_names)

stats_deseq <- stats_deseq[!is.na(stats_deseq)]
stats_deseq <- sort(stats_deseq, decreasing = TRUE)
                        
dex_names <- sub("\\.\\d+$", "", dex_gene$gene_id)
dex_vals <- dex_gene$stat
stats_dex <- setNames(dex_vals, dex_names)

stats_dex <- stats_dex[!is.na(stats_dex)]
stats_dex <- sort(stats_dex, decreasing = TRUE)
```

```{r}
fg_deseq <- fgseaMultilevel(pathways = BP_list, stats = stats_deseq,
                            minSize = 15, maxSize = 500, scoreType = "std")

dim(fg_deseq)
```

```{r}
fg_dex <- fgseaMultilevel(pathways = BP_list, stats = stats_dex,
                          minSize = 15, maxSize = 500, scoreType = "std")

dim(fg_dex)
```

```{r}
sig_deseq <- fg_deseq  |>  filter(padj < alpha)  |>  arrange(padj)
sig_dex <- fg_dex  |>  filter(padj < alpha) |>  arrange(padj)

n_sig_de <- nrow(sig_deseq)
n_sig_dx <- nrow(sig_dex)

# shared pathways (by name)
shared <- intersect(sig_deseq$pathway, sig_dex$pathway)
n_shared <- length(shared)

n_sig_de
n_sig_dx
n_shared
```

```{r}
shared
```

```{r}
de_df <- fg_deseq |> transmute(pathway, NES_de = NES, padj_de = padj)
dx_df <- fg_dex |> transmute(pathway, NES_dx = NES, padj_dx = padj)

# Merge on pathway (shared sets only)
cmp <- inner_join(de_df, dx_df, by = "pathway") %>%
  mutate(sig_cat = case_when(
    padj_de < alpha & padj_dx < alpha ~ "Both",
    padj_de < alpha & padj_dx >= alpha ~ "DESeq2 only",
    padj_de >= alpha & padj_dx < alpha ~ "DEXSeq only",
    TRUE ~ "None"
  )) %>%
  mutate(sig_cat = factor(sig_cat, levels = c("Both","DESeq2 only","DEXSeq only","None")))


cmp
```

```{r}
# Scatter + smooth
ggplot(cmp, aes(x = NES_de, y = NES_dx, color = sig_cat)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "NES (DESeq2 fgsea)",
    y = "NES (DEXSeq fgsea)",
    color = "Significance",
    title = "NES comparison: DESeq2 vs DEXSeq (fgseaMultilevel)"
  ) +
  theme_minimal()
```

```{r}
save(OcularResult, file="data/Ocular_dxdResult.RData")
```

# Spinal Motor Neurons

### Load Data

```{r}
isoCm_Spinal <- read.csv("data/SpinalIsoformCountMatrix.csv", row.names = 1)
Spinal_Metadata <- read.csv("data/SpinalSampleMetadata.csv", row.names = 1)
IsoformInfo <- read.csv("data/isoformInfo.csv")
```

```{r}
SpinalResult <- dexseq_analysis(isoCm_Spinal, Spinal_Metadata, IsoformInfo)
```

### Compare to Differential Expression:

```{r}
alpha <- 0.05  # example threshold

SpinalResultGene <- SpinalResult |> 
  as.data.frame() |> 
  as_tibble() |>
  filter(!is.na(pvalue)) |> 
  select(
    gene_id = groupID,
    isoform_id = featureID,
    stat,
    log2FC = log2fold_SOD1_C9ORF72,
    pvalue,
    padj
  ) |> 
  group_by(gene_id) |> 
  slice_min(
    order_by = pvalue,
    n = 1,
    with_ties = FALSE
  )  |> 
  ungroup()  |> 
  mutate(
    padj = p.adjust(pvalue, method = 'fdr')
  ) |>
  filter(!is.na(padj))
```

```{r}
dex_gene <- SpinalResultGene |> 
  mutate(gene_id,
         dex_padj = padj,
         dex_sig = !is.na(padj) & padj < alpha)
```

```{r}
load("data/Spinal_dseqResult.RData")
SpinalDesq <- res_spinal
SpinalDesq$padj <- as.numeric(SpinalDesq$padj)
```

```{r}
SpinalDesq <- SpinalDesq |>
  as.data.frame() |> 
  rownames_to_column("gene_id") |>
  mutate(deseq_sig = !is.na(padj) & padj < alpha) |>
  filter(!is.na(padj))
```

```{r}
# Gene universe = genes present in BOTH results
univ <- intersect(dex_gene$gene_id, SpinalDesq$gene_id)
both <- dex_gene %>%
  filter(gene_id %in% univ) %>%
  mutate(gene_id, dex_sig) %>%
  inner_join(SpinalDesq %>%
    filter(gene_id %in% univ) %>%
    mutate(gene_id, deseq_sig), by = "gene_id")

contig_tab <- table(DEXSeq = both$dex_sig, DESeq2 = both$deseq_sig)
contig_tab
```

**Fisher Test**

```{r}
ft <- fisher.test(contig_tab, alternative = "greater")
ft
```

### Compare to Differential Expression at Gene-set Level

```{r}
BP_df = msigdbr(species = "human", category = "C5", subcategory = "BP")
BP_list = split(BP_df$ensembl_gene, BP_df$gs_name)
```

```{r}
deseq_names <- sub("\\.\\d+$", "", SpinalDesq$gene_id)
deseq_vals <- SpinalDesq$stat
stats_deseq <- setNames(deseq_vals, deseq_names)

stats_deseq <- stats_deseq[!is.na(stats_deseq)]
stats_deseq <- sort(stats_deseq, decreasing = TRUE)
                        
dex_names <- sub("\\.\\d+$", "", dex_gene$gene_id)
dex_vals <- dex_gene$stat
stats_dex <- setNames(dex_vals, dex_names)

stats_dex <- stats_dex[!is.na(stats_dex)]
stats_dex <- sort(stats_dex, decreasing = TRUE)
```

```{r}
fg_deseq <- fgseaMultilevel(pathways = BP_list, stats = stats_deseq,
                            minSize = 15, maxSize = 500, scoreType = "std")


fg_dex <- fgseaMultilevel(pathways = BP_list, stats = stats_dex,
                          minSize = 15, maxSize = 500, scoreType = "std")
```

```{r}
sig_deseq <- fg_deseq %>% filter(padj < alpha) %>% arrange(padj)
sig_dex <- fg_dex %>% filter(padj < alpha) %>% arrange(padj)
n_sig_de <- nrow(sig_deseq)
n_sig_dx <- nrow(sig_dex)
# shared pathways (by name)
shared <- intersect(sig_deseq$pathway, sig_dex$pathway)
n_shared <- length(shared)

n_sig_de
n_sig_dx
n_shared
```

```{r}
shared
```

```{r}
de_df <- fg_deseq %>% transmute(pathway, NES_de = NES, padj_de = padj)
dx_df <- fg_dex %>% transmute(pathway, NES_dx = NES, padj_dx = padj)
# Merge on pathway (shared sets only)
cmp <- inner_join(de_df, dx_df, by = "pathway") %>%
  mutate(sig_cat = case_when(
    padj_de < alpha & padj_dx < alpha ~ "Both",
    padj_de < alpha & padj_dx >= alpha ~ "DESeq2 only",
    padj_de >= alpha & padj_dx < alpha ~ "DEXSeq only",
    TRUE ~ "None"
  )) %>%
  mutate(sig_cat = factor(sig_cat, levels = c("Both","DESeq2 only","DEXSeq only","None")))
```

```{r}
# Scatter + smooth
ggplot(cmp, aes(x = NES_de, y = NES_dx, color = sig_cat)) +
geom_point(alpha = 0.8) +
geom_smooth(method = "lm", se = FALSE) +
labs(
x = "NES (DESeq2 fgsea)",
y = "NES (DEXSeq fgsea)",
color = "Significance",
title = "NES comparison: DESeq2 vs DEXSeq (fgseaMultilevel)"
) +
theme_minimal()
```

```{r}
save(SpinalResult, file="data/Spinal_dxdResult.RData")
```

```{r}

iso_id_padj_value_ocular = data.frame("transcript_id" = OcularResult$featureID,
                                       "padj" = OcularResult$padj)

iso_id_padj_value_spinal = data.frame("transcript_id" = SpinalResult$featureID,
                                       "padj" = SpinalResult$padj)

write.csv(iso_id_padj_value_ocular, "data/iso_id_padj_value_ocular.csv")
write.csv(iso_id_padj_value_spinal, "data/iso_id_padj_value_spinal.csv")
```
